---
title: "BA 64060 Assignment 1"
author: "Madeline Witzeman"
date: "2023-09-10"
output:
  html_document: default
  pdf_document: default
---
## Step 1 & 2: Download dataset and Import dataset into R

```{r setup, include=FALSE}
ds <- read.csv("C:\\Users\\Madeline\\Documents\\Real_Estate_Sales_2001-2020_GL.csv")
```

## The code above assigns 'ds' as my dataset and imports the dataset into R from my Documents 
## The dataset was downloaded from https://catalog.data.gov/dataset/real-estate-sales-2001-2018

## Step 3: Print descriptive stats for quantitative and categorical variables

```{r}

min(ds$Assessed.Value, na.rm=TRUE)
max(ds$Assessed.Value, na.rm=TRUE)
range(ds$Assessed.Value, na.rm=TRUE)
mean(ds$Assessed.Value, na.rm=TRUE)
median(ds$Assessed.Value, na.rm=TRUE)
sd(ds$Assessed.Value, na.rm=TRUE)

min(ds$Sale.Amount, na.rm=TRUE)
max(ds$Sale.Amount, na.rm=TRUE)
range(ds$Sale.Amount, na.rm=TRUE)
mean(ds$Sale.Amount, na.rm=TRUE)
median(ds$Sale.Amount, na.rm=TRUE)
sd(ds$Sale.Amount, na.rm=TRUE)

dplyr::count(ds, Property.Type)
dplyr::count(ds, Residential.Type)

table(ds$Residential.Type)
tablepercent <- table(ds$Residential.Type)
prop.table(tablepercent)

```

## In the code above, I've chosen to print the min, max, range, mean, median, and standard deviation for the quantitative variables 'Assessed.Value' and 'Sale.Amount'
## I also printed the count for the categorical variables 'Property. Type' and 'Residential.Type' and printed proportions or precentages for 'Residential.Type'


## Step 4: Transform a variable

```{r}
ds = dplyr::rename(ds, Year.Listed = List.Year)
head(ds)

```

## In the code above, I renamed the variable 'List.Year' to 'Year.Listed' using the dplyr package


## Step 5: Plot a quantitative variable, including a scatterplot

```{r echo = FALSE}

plot(ds$Assessed.Value,ds$Sale.Amount, xlab = "Assessed Value ($)", ylab = "Sale Price ($)")

```

## The code above produces a scatterplot of 'Assessed.Value' and 'Sale.Amount'

```{r echo = FALSE}

SalePrice <- ds$Sale.Amount
hist(SalePrice, main="Property Sales up to $1M in CT Between 2001-2020", xlab = "Sale Price ($)", breaks = "Scott", xlim = c(1,1000000))

```

## The code above produces a histogram of 'Sale.Amount' with the x-axis limited to values between $1 and $1,000,000 and an optimized amount of bins

